---
title: "Agriculture Analysis"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(sf)
library(here)
library(mapview)
```

### Read in Data

[**`st_read()`**](https://r-spatial.github.io/sf/reference/st_read.html) reads a spatial file (here, GeoJSON) into an `sf` object — a data frame with a special geometry list-column attached. This is the entry point for everything that follows, so any problem here (wrong CRS, dropped features, malformed geometry) propagates through the whole analysis.

```{r}
ag_2015 <- st_read(here("data/Agricultural_Land_Use_2015_Baseline.geojson"))
ag_2020 <- st_read(here("data/Agricultural_Land_Use_2020_Update.geojson"))
```

### Check 1: Basic structural consistency between the two datasets

Before doing *any* analysis, verify the two files are actually comparable — same CRS, same geometry type, sane feature counts, consistent naming.

[**`st_crs()`**](https://r-spatial.github.io/sf/reference/st_crs.html) retrieves (or sets) an object's Coordinate Reference System. Comparing `st_crs(ag_2015) == st_crs(ag_2020)` catches a mismatch before it silently corrupts every distance or area calculation downstream — this is a pure inspection call, it doesn't change the data.

[**`st_geometry_type()`**](https://r-spatial.github.io/sf/reference/st_geometry_type.html) returns the geometry type (POLYGON, MULTIPOLYGON, etc.) per feature, or with `by_geometry = FALSE`, one summary value. Mixed types between years can break `rbind()` or comparisons that assume uniform geometry.

```{r}
# CRS must match, or every downstream operation is silently wrong
st_crs(ag_2015) == st_crs(ag_2020)

# Geometry type consistency
st_geometry_type(ag_2015, by_geometry = FALSE)
st_geometry_type(ag_2020, by_geometry = FALSE)

# Feature counts and column names — note cropcatego vs crops_2020 already differ
nrow(ag_2015); nrow(ag_2020)
names(ag_2015); names(ag_2020)

# Island naming must match exactly for any group_by(island) later —
# a stray "Big Island " vs "Big Island" or "Molokai" vs "Moloka'i" breaks joins silently
sort(unique(ag_2015$island))
sort(unique(ag_2020$island))
```

### Check 2: Geometry validity — how much are we actually fixing?

`st_make_valid()` is often called reflexively without checking what it's doing. Quantify it instead.

[**`st_is_valid()`**](https://r-spatial.github.io/sf/reference/valid.html) checks whether each geometry follows OGC simple-features validity rules (no self-intersections, properly closed rings, etc.), returning `TRUE`/`FALSE`/`NA` per feature. Counting invalid features *before* fixing anything turns an invisible cleanup step into a quantified, reportable fact.

[**`st_make_valid()`**](https://r-spatial.github.io/sf/reference/valid.html) — shares its documentation page with `st_is_valid()` — repairs invalid geometries (self-intersections, bowties, etc.) using GEOS or s2 depending on the CRS, without altering geometries that were already valid.

[**`st_area()`**](https://r-spatial.github.io/sf/reference/geos_measures.html) computes the area of each geometry as a `units` object, respecting whatever length unit the CRS uses. Measuring total area *before and after* `st_make_valid()` is the critical pairing here: a "repair" that also meaningfully shrinks or grows total area is a signal something more serious than topology noise is happening. A shift of a few dozen square meters against a statewide footprint of hundreds of millions of square meters is negligible; a shift of a meaningful percentage is not.

```{r}
invalid_2015 <- sum(!st_is_valid(ag_2015))
invalid_2020 <- sum(!st_is_valid(ag_2020))

invalid_2015; invalid_2020

# Total acreage before fixing, so we can confirm fixing didn't distort the data
area_before_2015 <- sum(st_area(ag_2015))
area_before_2020 <- sum(st_area(ag_2020))

ag_2015 <- st_make_valid(ag_2015)
ag_2020 <- st_make_valid(ag_2020)

area_after_2015 <- sum(st_area(ag_2015))
area_after_2020 <- sum(st_area(ag_2020))

# These differences should be TINY (topology fixes, not real area change).
# A large difference here means something is structurally wrong with the source data.
as.numeric(area_after_2015 - area_before_2015)
as.numeric(area_after_2020 - area_before_2020)
```

### Make a map

Fifteen unique `cropcatego` values is too many categories for most default qualitative palettes — `mapview`'s default and `RColorBrewer`'s largest qualitative sets top out well below that, and interpolating them past their design range produces muddy, hard-to-distinguish colors. `viridis`-style palettes are built for *ordered* data and can misleadingly imply a ranking across unordered categories like crop type, so they're a poor fit here too. The `pals` package provides palettes purpose-built for many discrete, unordered categories:

```{r}
#install.packages("pals")
library(pals)

n_cats <- length(unique(ag_2015$cropcatego))

mapview(ag_2015, zcol = "cropcatego",
        col.regions = alphabet(n_cats),
        layer.name = "Crop Category")
```

### Questions: Did we add/lose ag land acreage between 2015 and 2020?

```{r}
sum_2015_acreage <- sum(ag_2015$acreage)
sum_2020_acreage <- sum(ag_2020$acreage)

net_change_attribute <- sum_2020_acreage - sum_2015_acreage
net_change_attribute

```

### Check 3: Does the `acreage` attribute agree with the geometry itself?

The `acreage` column is a *stored* value, not something computed live from the polygon. If it was calculated in a different CRS, at a different time, or has data-entry errors, attribute-based stats above could be wrong even though the geometry itself is fine. Cross-checking a stored attribute against an independently computed value is a habit worth applying anywhere a dataset carries a number that claims to describe its own geometry.

[**`st_transform()`**](https://r-spatial.github.io/sf/reference/st_transform.html) reprojects geometry from one CRS to another — here, WGS84 lon/lat to UTM Zone 4N (EPSG:32604). This is necessary before any area, distance, or overlay operation, since GEOS-based planar operations are only numerically reliable in a projected CRS; doing them directly on geographic coordinates distorts results, especially across an archipelago spanning multiple degrees of latitude.

```{r}
ag_2015_proj <- st_transform(ag_2015, 32604)
ag_2020_proj <- st_transform(ag_2020, 32604)

ag_2015_proj <- ag_2015_proj %>%
  mutate(computed_acres = as.numeric(st_area(.)) / 4046.86,
         acre_diff = computed_acres - acreage,
         pct_diff = 100 * acre_diff / acreage)

# Ideally these are all close to 0. Flag anything with a large mismatch.
summary(ag_2015_proj$pct_diff)

ggplot(ag_2015_proj, aes(x = pct_diff)) +
  geom_histogram(bins = 50) +
  labs(title = "2015: % difference between stored acreage and geometry-computed acreage",
       x = "% difference", y = "Count of parcels")
```

[**`st_drop_geometry()`**](https://r-spatial.github.io/sf/reference/st_geometry.html) strips the geometry column, returning a plain data frame/tibble. Use it whenever doing ordinary `dplyr` summarizing (grouping, counting, `arrange()`) that doesn't need geometry.

```{r}
# Which specific parcels disagree most? Worth spot-checking these on the map.
ag_2015_proj %>%
  st_drop_geometry() %>%
  arrange(desc(abs(pct_diff))) %>%
  select(objectid, cropcatego, acreage, computed_acres, pct_diff) %>%
  head(10)
```

If most parcels cluster near 0% and a handful are wildly off, that's a signal of digitizing errors or unit mix-ups in specific records — not a systemic problem, but worth flagging or excluding before trusting totals built from `acreage`.

### Check 4: Are "rows" the same thing as "distinct shapes"?

A `MULTIPOLYGON` dataset can store more than one physically disconnected shape under a single row — one `objectid`/`cropcatego` record might represent several separate fields belonging to the same parcel entry. Before reasoning about parcel counts, it's worth confirming whether that's happening, since `nrow()` alone can understate how many actual distinct shapes are present.

```{r}
# How many individual polygon parts does each row actually contain?
parts_per_row <- sapply(st_geometry(ag_2015_proj), function(g) length(g[[1]]))

table(parts_per_row)   # rows with 1 part vs. rows bundling several parts

sum(parts_per_row)     # total distinct shapes across all rows, prior to any dissolving
```

A dataset where most rows have exactly one part, but a meaningful subset have several, tells you the "row count" and "shape count" are two different numbers worth quoting separately in any summary of parcel complexity.

### Which land was converted? (Spatial footprint method)

[**`st_union()`**](https://r-spatial.github.io/sf/reference/geos_combine.html) dissolves multiple geometries into a single combined footprint — here, collapsing every individual parcel for a year into one "footprint of all ag land that year" geometry. The result is a single `sfc` object (all attribute columns are dropped in the process), stored internally as one `MULTIPOLYGON` — a container that can hold many physically disconnected pieces, not necessarily one contiguous shape. Two parcels only get topologically fused into a single shared piece where they actually touch or overlap; everywhere else, they remain separate parts inside that one object.

```{r}
union_2015 <- st_union(ag_2015_proj)
union_2020 <- st_union(ag_2020_proj)
```

### Check 5: Did dissolving change total area?

If parcels within the same year genuinely overlap (share interior area, not just a boundary edge), the dissolved union will have *less* total area than the sum of the individual parcel areas, since overlapping regions only get counted once after dissolving. Comparing the two totals is a fast way to detect that before doing any heavier overlap diagnostics.

```{r}
sum_individual_2015 <- sum(as.numeric(st_area(ag_2015_proj)))
sum_dissolved_2015  <- as.numeric(st_area(union_2015))

sum_individual_2020 <- sum(as.numeric(st_area(ag_2020_proj)))
sum_dissolved_2020  <- as.numeric(st_area(union_2020))

# If these differ noticeably, some parcels overlap each other within the same year
sum_individual_2015 - sum_dissolved_2015
sum_individual_2020 - sum_dissolved_2020
```

A difference near zero (floating-point noise) means that year's parcels don't meaningfully overlap. A substantial difference — convert it to acres (`/ 4046.86`) for an interpretable number — means real double-counted area exists somewhere in that year's dataset, and is worth locating directly:

[**`st_overlaps()`**](https://r-spatial.github.io/sf/reference/geos_binary_pred.html) is the correct predicate for this: it returns `TRUE` only when two geometries share actual interior area, and specifically excludes geometries that merely touch along a shared edge (which is normal and expected for adjacent farm parcels). This distinction matters — a broader predicate like `st_intersects()` will also flag simple edge-sharing neighbors, producing a large, evenly distributed set of "flagged" parcels that don't actually represent double-counted area. `st_overlaps()` isolates genuine overlap specifically.

```{r}
real_overlaps <- st_overlaps(ag_2020_proj)
has_real_overlap <- lengths(real_overlaps) > 0

sum(has_real_overlap)

overlapping_parcels <- ag_2020_proj[has_real_overlap, ]

# Total acreage represented by the overlapping parcels themselves
overlapping_parcels %>%
  st_drop_geometry() %>%
  summarize(total_acreage = sum(acreage))

# Is it concentrated on particular islands or crop categories?
overlapping_parcels %>%
  st_drop_geometry() %>%
  count(island, sort = TRUE)

overlapping_parcels %>%
  st_drop_geometry() %>%
  count(crops_2020, sort = TRUE)
```

A useful cross-check: the acreage measured via the union-area-difference above should be in the same ballpark as the overlap found through `st_overlaps()` — if the two numbers are wildly different in scale, that's worth investigating before drawing conclusions from either one alone.

[**`st_join()`**](https://r-spatial.github.io/sf/reference/st_join.html) and direct visual inspection help characterize *what kind* of overlap is present once flagged parcels are identified:

```{r}
# Inspect one overlapping cluster directly
example_id <- which(has_real_overlap)[1]
neighbors <- real_overlaps[[example_id]]
cluster <- ag_2020_proj[c(example_id, neighbors), ]

cluster %>% st_drop_geometry() %>% select(objectid, crops_2020, acreage)

mapview(cluster, zcol = "crops_2020")
```

Common patterns worth distinguishing: duplicate digitizing of the same field, a boundary redraw where a field was split between two categories without being cleanly trimmed against its neighbor, or a genuine data-entry conflict. Each implies a different fix — deduplication, clipping to a priority order, or exclusion with a documented rationale.

### Building the change map

[**`st_difference()`**](https://r-spatial.github.io/sf/reference/geos_binary_ops.html) returns the parts of geometry `x` that do *not* overlap with geometry `y`. `st_difference(union_2020, union_2015)` gives land present in 2020 but not 2015 ("new"); reversing the arguments gives the reverse ("lost"). This function and `st_intersection()` share one documentation page, since they're two sides of the same underlying set operation — argument order controls direction, so it's worth verifying which side is which rather than assuming from memory.

[**`st_sf()`**](https://r-spatial.github.io/sf/reference/sf.html) constructs an `sf` object from attribute columns plus a geometry column — needed here because `st_difference()` alone returns a bare geometry with no attributes attached.

[**`st_sfc()`**](https://r-spatial.github.io/sf/reference/sfc.html) creates a geometry list-column and lets you (re)assign a CRS via `crs =`, since `st_difference()` on unioned geometries returns a geometry object without CRS metadata attached.

[**`st_cast()`**](https://r-spatial.github.io/sf/reference/st_cast.html) converts geometries from one type to another — here, normalizing a mixed `GEOMETRYCOLLECTION` (which `st_difference()` can produce at sliver edges) into clean `MULTIPOLYGON` for plotting.

```{r}
# New ag land: in 2020 but NOT in 2015
new_2020 <- st_difference(union_2020, union_2015)

# Lost ag land: in 2015 but NOT in 2020
lost_2015 <- st_difference(union_2015, union_2020)

new_sf  <- st_sf(status = "New (2020 only)", geometry = st_sfc(new_2020, crs = 32604))
lost_sf <- st_sf(status = "Lost (2015 only)", geometry = st_sfc(lost_2015, crs = 32604))

changes <- rbind(
  st_cast(new_sf, "MULTIPOLYGON"),
  st_cast(lost_sf, "MULTIPOLYGON")
) %>%
  st_transform(4326)
```

### Check 6: Sliver inspection — is "change" real or just boundary noise?

Two independently digitized surveys almost never line up perfectly at parcel edges. Before trusting `changes`, look at the size distribution of individual change-polygons — a sea of tiny slivers means most of the "new/lost" area is boundary noise, not real land-use change.

`st_cast(..., "POLYGON")` is used here for a different purpose than above: it *splits* multipart geometries into individual single-polygon rows, so each disconnected sliver or parcel becomes its own measurable row. A dissolved multipolygon would hide the sliver problem if left uncast.

```{r}
changes_pieces <- changes %>%
  st_transform(32604) %>%
  st_cast("POLYGON") %>%
  mutate(piece_acres = as.numeric(st_area(.)) / 4046.86)

nrow(changes_pieces)

ggplot(changes_pieces, aes(x = piece_acres)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  facet_wrap(~status) +
  labs(title = "Distribution of individual change-polygon sizes (log scale)",
       x = "Acres (log scale)", y = "Count")

# How much total area is contributed by tiny slivers vs. real parcels?
changes_pieces %>%
  st_drop_geometry() %>%
  mutate(size_class = if_else(piece_acres < 1, "Sliver (<1 acre)", "Substantial (>=1 acre)")) %>%
  group_by(status, size_class) %>%
  summarize(n = n(), total_acres = sum(piece_acres), .groups = "drop")
```

If slivers under 1 acre make up hundreds of polygons but a tiny fraction of total acreage, filtering them out is safe and mostly cosmetic. If they make up a large share of total acreage, that's worth stating explicitly rather than silently filtering it away.

```{r}
changes_clean <- changes_pieces %>%
  filter(piece_acres >= 1) %>%
  st_transform(4326)

# Confirm how much area (and how many features) the filter removed
removed <- nrow(changes_pieces) - nrow(changes_clean)
removed_acres <- sum(changes_pieces$piece_acres) - sum(changes_clean$piece_acres)
removed; removed_acres
```

```{r}
mapview(changes_clean, zcol = "status",
        col.regions = c("Lost (2015 only)" = "firebrick",
                        "New (2020 only)" = "forestgreen"),
        layer.name = "Ag Land Change 2015–2020")
```

### Check 7: The capstone check — do two independent methods agree?

There are now two completely different ways of measuring net ag land change:

1.  **Attribute method**: sum(`acreage` in 2020) − sum(`acreage` in 2015)
2.  **Geometry method**: area(new_2020) − area(lost_2015), from spatial overlay

These used different data (a stored column vs. computed geometry) and different logic (simple subtraction vs. polygon differencing) — if they roughly agree, that's strong evidence the pipeline is sound. If they diverge, the earlier checks (multi-part geometry, overlapping parcels, sliver noise, acreage-vs-geometry mismatch) are the toolkit for isolating why, rather than treating the discrepancy as unexplainable.

```{r}
net_change_geometric <- (as.numeric(st_area(new_2020)) - as.numeric(st_area(lost_2015))) / 4046.86

comparison <- tibble(
  method = c("Attribute-based (sum of acreage column)",
             "Geometry-based (spatial overlay of dissolved footprints)"),
  net_change_acres = c(net_change_attribute, net_change_geometric)
)

comparison

gap_acres <- diff(comparison$net_change_acres)
gap_pct <- 100 * gap_acres / mean(comparison$net_change_acres)

gap_acres; gap_pct
```

**Discussion prompts for class**, once students get their own gap value:

- If the two methods agree closely, what does that imply about data quality across the two survey years?
- If they diverge, is the gap explained by overlapping parcels (Check 5), sliver noise (Check 6), the acreage-vs-geometry mismatch (Check 3), or something else? Walk through the checks in order to isolate it rather than guessing.
- A clean number is not automatically a meaningful one — always ask "changed from what, to what" before reporting a raw net change, since a category disappearing from one year's schema can produce an apparent "loss" that isn't a real land-use change.

### Category schema consistency check

```{r}
# Are the same crop categories even present across both years?
cats_2015 <- unique(ag_2015$cropcatego)
cats_2020 <- unique(ag_2020$crops_2020)

setdiff(cats_2015, cats_2020)  # categories that vanished after 2015
setdiff(cats_2020, cats_2015)  # categories that appeared new in 2020
```

A non-empty result here means part of any acreage "loss" calculated above might be a **classification change**, not a **land-use change** — this is a key sanity check before drawing conclusions from this dataset.

------------------------------------------------------------------------

### Function reference summary

| Function | Purpose in this analysis | Documentation |
|----|----|----|
| `st_read()` | Load spatial files into `sf` objects | [link](https://r-spatial.github.io/sf/reference/st_read.html) |
| `st_crs()` | Inspect/compare coordinate reference systems | [link](https://r-spatial.github.io/sf/reference/st_crs.html) |
| `st_geometry_type()` | Check geometry type consistency | [link](https://r-spatial.github.io/sf/reference/st_geometry_type.html) |
| `st_is_valid()` | Detect invalid geometries | [link](https://r-spatial.github.io/sf/reference/valid.html) |
| `st_make_valid()` | Repair invalid geometries | [link](https://r-spatial.github.io/sf/reference/valid.html) |
| `st_area()` | Compute polygon area | [link](https://r-spatial.github.io/sf/reference/geos_measures.html) |
| `st_transform()` | Reproject to a different CRS | [link](https://r-spatial.github.io/sf/reference/st_transform.html) |
| `st_union()` | Dissolve/merge geometries | [link](https://r-spatial.github.io/sf/reference/geos_combine.html) |
| `st_difference()` | Subtract one geometry set from another | [link](https://r-spatial.github.io/sf/reference/geos_binary_ops.html) |
| `st_intersection()` | Find overlapping area between geometry sets | [link](https://r-spatial.github.io/sf/reference/geos_binary_ops.html) |
| `st_overlaps()` | Test for true interior-area overlap, excluding simple touching | [link](https://r-spatial.github.io/sf/reference/geos_binary_pred.html) |
| `st_intersects()` | Test for any spatial contact (touching or overlapping); useful as a fast candidate pre-filter, not for isolating true overlap | [link](https://r-spatial.github.io/sf/reference/geos_binary_pred.html) |
| `st_sf()` | Build an `sf` object from attributes + geometry | [link](https://r-spatial.github.io/sf/reference/sf.html) |
| `st_sfc()` | Build/re-CRS a geometry list-column | [link](https://r-spatial.github.io/sf/reference/sfc.html) |
| `st_cast()` | Convert between geometry types (e.g., collection → multipolygon, or multipolygon → polygon) | [link](https://r-spatial.github.io/sf/reference/st_cast.html) |
| `st_drop_geometry()` | Strip geometry for plain data-frame operations | [link](https://r-spatial.github.io/sf/reference/st_geometry.html) |
| `st_join()` | Spatial (not attribute-key) join between two `sf` objects | [link](https://r-spatial.github.io/sf/reference/st_join.html) |

### Your Turn

I showed you where the change was, what was the ag land designation (cropcatego) with the most significant acreage lost\

```{r}
cat_2015 <- tapply(ag_2015$acreage, ag_2015$cropcatego, sum)
cat_2020 <- tapply(ag_2020$acreage, ag_2020$crops_2020, sum)

cats <- union(names(cat_2015), names(cat_2020))

change <- cat_2020[cats] - cat_2015[cats]
names(change) <- cats
change[is.na(change)] <- 0

sort(change)
```

```{r}
df_change <- data.frame(crop = names(change), change = as.numeric(change))

ggplot(df_change, aes(x = reorder(crop, change), y = change)) +
  geom_col() +
  coord_flip() +
  xlab("Crops") +
  ylab("Value")
```

I tried to add the colors to show plus or minus.
