---
title: "Isochrones"
format: html
editor: visual
---

An **isochrone** is a line or polygon on a map that connects all points reachable from a specific starting location within a set amount of time

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

### Libraries

```{r}
library(mapboxapi)
library(leaflet)
library(mapview)
```

### Create Isochrone

```{r}
isos <- mb_isochrone(
  location = "One Bowerman Dr, Beaverton, OR 97005",
  profile = "driving",
  time = 1:45
)
```

```{r}
mapview(isos, zcol = "time")
```

```{r}
library(sf)
library(fasterize)

isos_proj <- st_transform(isos, 32618)

template <- raster(isos_proj, resolution = 100)

iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

iso_surface
pal <- colorNumeric("plasma", isos$time, na.color = "transparent")

nike_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time from<br>Nike HQ")

nike_map
```

```{r}
nike_icon <- makeIcon(
  iconUrl = "https://nike.com/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

nike_hq <- mb_geocode("One Bowerman Dr, Beaverton, OR 97005")

nike_map %>%
  addMarkers(lng = nike_hq[1], lat = nike_hq[2], icon = nike_icon)
```

### Your turn - Recreate for Chaminade

```{r}

```

```{r}

```

### Lets go further 

Question: Can we make an accessibility surfaces map for trauma centers, ER's, hospitals, fire stations etc. Maybe combine all of them into one shiny app. Let's strategise and design...
