---
title: "Isochrones"
format: html
editor: visual
---

An **isochrone** is a line or polygon on a map that connects all points reachable from a specific starting location within a set amount of time

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

### Libraries

```{r}
#install.packages('mapboxapi')
library(mapboxapi)
library(leaflet)
library(mapview)

```

### Create Isochrone

```{r}
isos <- mb_isochrone(
  location = "One Bowerman Dr, Beaverton, OR 97005",
  profile = "driving-traffic",
  time = 1:45,
  depart_at = "2026-10-01T16:00"
)
```

```{r}
mapview(isos, zcol = "time")
```

```{r}
library(sf)
library(fasterize)


isos_proj <- st_transform(isos, 32618)

template <- raster(isos_proj, resolution = 100)

iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

iso_surface
pal <- colorNumeric("plasma", isos$time, na.color = "transparent")

nike_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time from<br>Nike HQ")

nike_map
```

```{r}
nike_icon <- makeIcon(
  iconUrl = "https://nike.com/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

nike_hq <- mb_geocode("One Bowerman Dr, Beaverton, OR 97005")

nike_map %>%
  addMarkers(lng = nike_hq[1], lat = nike_hq[2], icon = nike_icon)
```

### Your turn - Recreate for Chaminade

```{r}
chaminade_isos <- mb_isochrone(
  location = "3140 Waialae Ave, Honolulu, HI 96816",
  profile = "driving",
  time = 1:45 
)
```

```{r}
mapview(chaminade_isos, zcol = "time")
```

```{r}
chaminade_proj <- st_transform(chaminade_isos, 32604)

chaminade_template <- raster(chaminade_proj, resolution = 100)

chaminade_surface <- fasterize(chaminade_proj, chaminade_template,
                               field = "time", fun = "min")

chaminade_pal <- colorNumeric("plasma", chaminade_isos$time,
                              na.color = "transparent")
chaminade_pal

chaminade_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(chaminade_surface, colors = chaminade_pal, opacity = 0.5) %>%
  addLegend(values = chaminade_isos$time, pal = chaminade_pal,
            title = "Drive-time from<br>Chaminade University")

chaminade_map
```

```{r}
chaminade_icon <- makeIcon(
  iconUrl = "https://ucm.sfo3.digitaloceanspaces.com/wp-content/uploads/2018/08/08074024/Chaminade-Seal.png",
  iconWidth = 25,
  iconHeight = 25
)

chaminade_hq <- mb_geocode("3140 Waialae Ave, Honolulu, HI 96816")

chaminade_map %>%
  addMarkers(lng = chaminade_hq[1], lat = chaminade_hq[2], icon = chaminade_icon)
```

### Lets go further

Question: Can we make an accessibility surfaces map for trauma centers, ER's, hospitals, fire stations etc. Maybe combine all of them into one shiny app. Let's strategise and design...

App Design

fire station :'mb_isochrone' works well \<- I gonna focus on.

amburance:

shelter:

can't go off base travel time in island wide emergency situation

hospital, trauma centers. etc...

senior transportation

police stations, hostage

swim, lifeguard response
