---
title: "nighttime_lights.qmd"
format: html
editor: visual
---

## **Installation** 

```{r}
# install.packages("devtools")
devtools::install_github("worldbank/blackmarbler")
```

##  Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
# Load packages
library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
library(lubridate)

#### Define NASA bearer token
bearer <- "eyJ0eXAiOiJKV1QiLCJvcmlnaW4iOiJFYXJ0aGRhdGEgTG9naW4iLCJzaWciOiJlZGxqd3RwdWJrZXlfb3BzIiwiYWxnIjoiUlMyNTYifQ.eyJ0eXBlIjoiVXNlciIsInVpZCI6InJlaW1uMTciLCJleHAiOjE3OTI4Nzk4MDgsImlhdCI6MTc4NzY5NTgwOCwiaXNzIjoiaHR0cHM6Ly91cnMuZWFydGhkYXRhLm5hc2EuZ292IiwiaWRlbnRpdHlfcHJvdmlkZXIiOiJlZGxfb3BzIiwiYWNyIjoiZWRsIiwiYXNzdXJhbmNlX2xldmVsIjozfQ.RbGg1fsSh6f3RsrcRuVkSKOjNlPePsjwyqfSztNOfKWeUPE3tBw3eD0ENb0raiUq4tctl2Rnn7Ny1OBPYwubr_3qHeiXodARL8nZmbr28OPDAbnRiNyyx0vBt3D8FZEO5M_qEfGxS-Y8j9bikrCOOxaEekNOsNbKTfSVg4aVHSCNWT0cOWkO4_Ne1Vkl-JNrPdjRW6seyNswEISS1Guluo6VYlPsI_ohLxYRRim9Ufjn-nXAWVEIxnBIUqt-nw7RliqUiscG0j8ZVChpRfD5ft6cbAF5V2-0Og710sWZSF-Nc_pykrs1u6uFPp4qS0HFGzpuSW_Y_MRXzjGLCXbAUQ"

### ROI
# Define region of interest (roi). The roi must be (1) an sf polygon and (2)
# in the WGS84 (epsg:4326) coordinate reference system. Here, we use the
# getData function to load a polygon of Ghana
roi_sf <- gadm(country = "GHA", level=1, path = tempdir()) 

####################################################
#| eval: false 

options(tigris_use_cache = TRUE)
library(tigris)
roi_hawaii <- counties(state = 'HI', cb = TRUE, year = 2022)
```

```{r}
#| eval: falase
library(terra)
crs(roi_hawaii)

# ID["EPSG/",4269]]"
```

need to match the ID

```{r}
#| eval: false
library(sf)
roi_hawaii <- roi_hawaii |> st_transform(4326) |> 
  st_crop(xmin = -160.5, xmax = -154.5, ymin = 18.5, ymax = 22.5)
crs(roi_hawaii)
```

```{r}
#| eval:false
library(blackmarbler)

bearer <- get_nasa_token(username = "reimn17", 
                         password = "C534dpcry/6b/#/")

r <- bm_raster(roi_sf = roi_hawaii,
               product_id = "VNP46A3", # VNP46A2 request the montly ,VNP46A2 is daily
               date = "2025-10-01",
               bearer = bearer)
```

```{r}
bearer <- get_nasa_token(username = "reimn17", 
                         password = "C534dpcry/6b/#/")
```

```{r}
# Preparing data 
#| eval:false
library(terra)

r <- r |> terra::mask(roi_hawaii)
r[] <- log(r[] + 1)
```

### **Make raster of nighttime lights**

```{r}
### Daily data: raster for February 5, 2021
r_20210205 <- bm_raster(roi_sf = roi_sf,
                        product_id = "VNP46A2",
                        date = "2021-02-05",
                        bearer = bearer)

### Monthly data: raster for October 2021
r_202110 <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A3",
                      date = "2021-10-01", # The day is ignored
                      bearer = bearer)

### Annual data: raster for 2021
r_2021 <- bm_raster(roi_sf = roi_sf,
                    product_id = "VNP46A4",
                    date = 2021,
                    bearer = bearer)
```

### **Make raster of nighttime lights across multiple time periods** 

```{r}
#### Daily data in March 2021
r_daily <- bm_raster(roi_sf = roi_sf,
                     product_id = "VNP46A2",
                     date = seq.Date(from = ymd("2021-03-01"), to = ymd("2021-03-31"), by = "day"),
                     bearer = bearer)

#### Monthly aggregated data in 2021 and 2022
r_monthly <- bm_raster(roi_sf = roi_sf,
                       product_id = "VNP46A3",
                       date = seq.Date(from = ymd("2021-01-01"), to = ymd("2022-12-01"), by = "month"),
                       bearer = bearer)

#### Yearly aggregated data in 2012 and 2021
r_annual <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A4",
                      date = 2012:2021,
                      bearer = bearer)
```

**Map of nighttime lights**

Using one of the rasters, we can make a map of nighttime lights

```{r}
#### Make raster
r <- bm_raster(roi_sf = roi_sf,
               product_id = "VNP46A3",
               date = "2021-10-01",
               bearer = bearer)

#### Prep data
r <- r |> terra::mask(roi_sf)

## Distribution is skewed, so log
r[] <- log(r[] + 1)

##### Map
ggplot() +
  geom_spatraster(data = r) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  labs(title = "Nighttime Lights: October 2021") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "none")
```

### **Trends over time**

```{r}
#### Extract annual data
ntl_df <- bm_extract(roi_sf = roi_sf,
                     product_id = "VNP46A4",
                     date = 2012:2022,
                     bearer = bearer)


```

```{r}
#### Trends over time
ntl_df |>
  ggplot() +
  geom_col(aes(x = date,
               y = ntl_sum),
           fill = "darkorange") +
  facet_wrap(~NAME_1) +
  labs(x = NULL,
       y = "NTL Luminosity",
       title = "Ghana Admin Level 1: Annual Average Nighttime Lights") +
  scale_x_continuous(labels = seq(2012, 2022, 4),
                     breaks = seq(2012, 2022, 4)) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

### **Workflow to update data**

```{r}
# Create directories to store data
dir.create(file.path(getwd(), "bm_files"))
dir.create(file.path(getwd(), "bm_files", "daily"))

# Extract daily-level nighttime lights data for Ghana's first administrative divisions.
# Save a separate dataset for each date in the `"~/Desktop/bm_files/daily"` directory.
# The code extracts data from January 1, 2023 to today. Given that daily nighttime lights
# data is produced on roughly a week delay, the function will only extract data that exists;
# it will skip extracting data for dates where data has not yet been produced by NASA Black Marble.
bm_extract(roi_sf = roi_sf,
           product_id = "VNP46A2",
           date = seq.Date(from = ymd("2023-01-01"), to = Sys.Date(), by = 1),
           bearer = bearer,
           output_location_type = "file",
           file_dir = file.path(getwd(), "bm_files", "daily"))

# Append daily-level datasets into one file
file.path(getwd(), "bm_files", "daily") |>
  list.files(pattern = "*.Rds",
             full.names = T) |>
  map_df(readRDS) |>
  saveRDS(file.path(getwd(), "bm_files", "ntl_daily.Rds"))Workflow to update data
```
