---
title: "Detecting a Power Outage from Space"
subtitle: "EAGLE-I outage reports + NASA Black Marble nighttime lights, Oahu, January 2025"
format: html
editor: visual
---

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(tigris)
library(sf)
library(blackmarbler)
library(purrr)
library(terra)
library(tidyterra)
library(lubridate)
library(arrow)
library(here)
library(exactextractr)
library(tmap)
library(mapview)
library(gt)
library(vembedr)

options(tigris_use_cache = TRUE)
```

# Summary

Hurricane Lala passed south of the Hawaiian Islands in mid-August 2026, striking Hawaiʻi Island as a Category 1 storm on August 15 before tracking west toward Oʻahu. Hawaiian Electric reported roughly 9,500 Oʻahu customers without power late on August 15; by 7 a.m. on August 16 that figure had risen to approximately 110,700, close to a quarter of the island's customer base.

**The question:** if a quarter of Oʻahu's electrical customers lose power overnight, can a satellite detect it from orbit?

**The answer:** yes, but faintly. Across the eight storm and recovery nights, 69% of Oʻahu's lit pixels were darker than baseline, compared with a measured baseline rate of 57% across seven pre-storm nights. The shift is consistent — every event night exceeded the baseline average — but modest. The most persuasive evidence is geographic: ranking Oʻahu communities by z-score independently surfaced the same three regions Hawaiian Electric named in its outage reports.

# Why nighttime lights can detect an outage

The Visible Infrared Imaging Radiometer Suite (VIIRS) aboard Suomi-NPP passes over Hawaiʻi at roughly 1–2 a.m. local time. NASA's Black Marble VNP46A2 product corrects that raw signal for moonlight, atmospheric effects, and viewing angle, producing a daily estimate of how much light each 500-metre pixel emits toward space.

Electric light is the dominant source of that signal over developed land. If the grid fails, the light goes with it — and because the correction removes the largest natural confounds, a night-to-night drop becomes interpretable as a change on the ground rather than a change in viewing conditions.

## Study design

Three periods structure the analysis:

| Period | Dates | Nights | Treatment |
|-----------------|-----------------|-----------------|-----------------------|
| Baseline | Aug 3 – Aug 13 | 11 requested, 7 returned | Averaged into one reference image |
| Storm | Aug 14 – Aug 17 | 4 | Compared individually |
| Recovery | Aug 18 – Aug 21 | 4 | Compared individually |

The baseline collapses to a per-pixel mean and standard deviation. Every storm and recovery night is compared against that same fixed reference, so all eight event nights are directly comparable to one another.

August 17 is placed in the storm window rather than recovery. VIIRS captures around 1 a.m., and the Oʻahu outage peaked overnight on August 16 into August 17, so that night is part of the event rather than the aftermath.

# Data

## Credentials

```{r bearer-token}

bearer <- get_nasa_token(username = "reimn17", 
                         password = "C534dpcry/6b/#/")
```

## Setting the baseline and storm dates

```{r set-dates}
# from 2026-08-03 to 2026-08-13 for baseline - average
# from 2026-08-14 to 2026-08-16 for storm - day by day
# form 2026-08-17 to 2026-08-21 for after - day by day
baseline_dates <- seq.Date(as.Date("2026-08-03"), as.Date("2026-08-13"), by = "day")
storm_dates    <- seq.Date(as.Date("2026-08-14"), as.Date("2026-08-17"), by = "day")
recovery_dates <- seq.Date(as.Date("2026-08-18"), as.Date("2026-08-21"), by = "day")

all_dates <- c(baseline_dates, storm_dates, recovery_dates)

period_lookup <- tibble(
  date = all_dates,
  period = factor(
    c(rep("Baseline", length(baseline_dates)),
      rep("Storm",    length(storm_dates)),
      rep("Recovery", length(recovery_dates))),
    levels = c("Baseline", "Storm", "Recovery")
  )
)

period_colors <- c(Baseline = "grey70", Storm = "firebrick", Recovery = "steelblue")
storm_window_label <- "Hurricane Lala, Aug 14-17, 2026"
```

## Study area boundary

`tigris` returns county boundaries in NAD83; Black Marble tiles are in WGS84, so a reprojection is required. The bounding box drops the Northwestern Hawaiian Islands, which Honolulu County technically includes but which contain no relevant infrastructure

```{r boundary}
roi_hawaii <- counties(state = "HI", cb = TRUE, year = 2022) |>
  st_transform(4326) |>
  st_crop(xmin = -160.5, xmax = -154.5, ymin = 18.5, ymax = 22.5) |>
  filter(NAMELSAD != "Kalawao County") |>
  mutate(county = gsub(" County", "", NAME))
```

## Pulling Black Marble rasters

Each date is requested individually inside `tryCatch()`, so a missing or failed date does not halt the batch. Layer names are explicitly overwritten with ISO dates, because `bm_raster()` returns names in a `t2026_08_14` format that later date matching cannot parse.

```{r pull-rasters}
pull_one_date <- function(d) {
  tryCatch({
    bm_raster(roi_sf = roi_hawaii,
              product_id = "VNP46A2",
              date = d,
              bearer = bearer,
              quality_flag_rm = 2)
  }, error = function(e) {
    message("Failed for ", d, ": ", conditionMessage(e))
    NULL
  })
}

r_list <- map(all_dates, pull_one_date)
names(r_list) <- as.character(all_dates)
r_list <- compact(r_list)

message("Dates successfully pulled: ", paste(names(r_list), collapse = ", "))
message("Dates dropped (no imagery / failed): ",
        paste(setdiff(as.character(all_dates), names(r_list)), collapse = ", "))

r_all <- rast(r_list)
names(r_all) <- names(r_list)
# Combine 

# Rec
baseline_ok <- intersect(as.character(baseline_dates), names(r_all))
storm_ok <- intersect(as.character(storm_dates), names(r_all))
recovery_ok <- intersect(as.character(recovery_dates), names(r_all))


# Rec
r_baseline <- r_all[[baseline_ok]]
r_storm <- r_all[[storm_ok]]
r_recovery <- r_all[[recovery_ok]]
```

#### **August 3 through 6 returned no imagery.** The baseline is therefore built from seven nights (August 7–13) rather than the eleven originally planned. All event nights were retrieved successfully.

## Data completeness

Cloud cover is the primary threat to this analysis. A storm night with few valid pixels could appear dark simply because the observation is missing, not because the lights went out. This check is the honesty test for everything that follows.

```{r mask-completeness}
# add recovery ver.
r_baseline <- terra::mask(r_baseline, roi_hawaii)
r_storm <- terra::mask(r_storm, roi_hawaii)
r_recovery <- terra::mask(r_recovery, roi_hawaii)

completeness_check <- tibble(
  date = c(baseline_ok, storm_ok, recovery_ok), #Rec
  period = c(rep("Baseline", length(baseline_ok)), rep("Storm", length(storm_ok)), rep("Recovery", length(recovery_ok))), #Rec
  n_valid_pixels = c(
    sapply(seq_len(nlyr(r_baseline)), function(i) sum(!is.na(values(r_baseline[[i]])))),
    sapply(seq_len(nlyr(r_storm)), function(i) sum(!is.na(values(r_storm[[i]])))),
    sapply(seq_len(nlyr(r_recovery)), function(i) sum(!is.na(values(r_recovery[[i]])))) # Rec
  )
)
print(completeness_check)
```

## Baseline versus event radiance, at four granularities

## Statewide and by county

County-level daily means are computed from the rasters already retrieved, rather than requesting the same tiles again through a separate summary function.

```{r county-extract}
extract_from_local <- function(raster_stack, roi_sf) {
  map_dfr(seq_len(nlyr(raster_stack)), function(i) {
    layer <- raster_stack[[i]]
    vals <- exact_extract(layer, roi_sf, fun = "mean")
    tibble(
      county = roi_sf$county,
      date = as.Date(names(raster_stack)[i]),
      ntl_mean = vals
    )
  })
}

ntl_trend <- bind_rows(
  extract_from_local(r_baseline, roi_hawaii) |> mutate(period = "Baseline"),
  extract_from_local(r_storm, roi_hawaii) |> mutate(period = "Storm"),
  extract_from_local(r_recovery, roi_hawaii) |> mutate(period = "Recovery")
) #Rec

ntl_trend <- ntl_trend |>
  mutate(period = factor(period, levels = c("Baseline", "Storm", "Recovery")))

statewide_trend <- ntl_trend |>
  group_by(date, period) |>
  summarize(ntl_mean = mean(ntl_mean, na.rm = TRUE), .groups = "drop")

ggplot(statewide_trend, aes(x = date, y = ntl_mean, fill = period)) +
  geom_col() +
  scale_fill_manual(values = c("Baseline" = "grey70", "Storm" = "firebrick", "Recovery" = "lightgreen")) + #Rec
  labs(title = "Statewide Mean Nighttime Radiance by Day",
       subtitle = "Baseline (Aug 7-13) vs. Storm (Aug 14-17) vs. Recovery (Aug 18-21), 2026",
       x = NULL, y = "Mean radiance", fill = NULL) +
  theme_minimal()
```

Broken out by county, Hawaiʻi County shows the deepest storm-period decline. That is consistent with the storm track: Hawaiʻi Island took the Category 1 landfall, while Oʻahu caught the weakening back end a day later. The contrast is useful context for why the Oʻahu signal is comparatively subtle.

```{r county-facet}

ggplot(ntl_trend, aes(x = date, y = ntl_mean, fill = period)) +
  geom_col() +
  facet_wrap(~county, scales = "free_y") +
  scale_fill_manual(values = c("Baseline" = "grey70", "Storm" = "firebrick", "Recovery" = "lightgreen")) +
  scale_x_date(date_breaks = "4 days", date_labels = "%b %d") +
  labs(title = "Mean Nighttime Radiance by Day, by County",
       subtitle = "Hawaiʻi County took the direct hit; Honolulu County is Oʻahu",
       x = NULL, y = "Mean radiance", fill = NULL) +
  theme_minimal()



```

## Scoping to Oʻahu

Hawaiian Electric reported the largest county-level customer impact on Oʻahu. From this point forward the pixel-level work is scoped to Honolulu County, which is the island of Oʻahu.

```{r scope-oahu}
oahu_boundary <- roi_hawaii |> filter(county == "Honolulu")

baseline_mean <- app(r_baseline, mean, na.rm = TRUE) |>
  crop(oahu_boundary) |> mask(oahu_boundary)
baseline_sd <- app(r_baseline, sd, na.rm = TRUE) |>
  crop(oahu_boundary) |> mask(oahu_boundary)
names(baseline_mean) <- "baseline_mean"
names(baseline_sd)   <- "baseline_sd"

storm_stack    <- r_storm    |> crop(oahu_boundary) |> mask(oahu_boundary)
recovery_stack <- r_recovery |> crop(oahu_boundary) |> mask(oahu_boundary)

# Storm and recovery nights as one stack — this is what the day-by-day
# maps facet over, and what replaces the old single-layer `storm_mean`.
event_stack <- c(storm_stack, recovery_stack)
```

## Pixel level: storm night versus recovery night

Every pixel has both a baseline value and a value on each event night. Plotting them against one another directly is a more informative comparison than examining either distribution alone. Points below the dashed identity line are darker than normal.

```{r paired-scatter}
peak_storm <- "2026-08-15"   # darkest night — check your completeness table
last_night <- "2026-08-21"   # end of recovery window

paired_values <- tibble(
  baseline = values(baseline_mean)[, 1],
  storm    = values(storm_stack[[peak_storm]])[, 1],
  recovery = values(recovery_stack[[last_night]])[, 1]
) |>
  filter(!is.na(baseline), !is.na(storm), !is.na(recovery))

paired_values |>
  pivot_longer(c(storm, recovery), names_to = "phase", values_to = "radiance") |>
  mutate(phase = factor(phase, levels = c("storm", "recovery"),
                        labels = c(paste("Storm:", peak_storm),
                                   paste("Recovered:", last_night)))) |>
  ggplot(aes(x = baseline, y = radiance)) +
  geom_point(alpha = 0.12, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linetype = "dashed") +
  facet_wrap(~phase) +
  labs(title = "Oʻahu: Per-Pixel Radiance vs. Baseline",
       subtitle = paste0(storm_window_label,
                         " — points below the line are darker than normal"),
       x = "Baseline radiance", y = "Night radiance") +
  theme_minimal()
```

```{r darkened-count}
paired_values |>
  summarize(
    n_total              = n(),
    pct_darkened_storm   = round(mean(storm    < baseline) * 100, 1),
    pct_darkened_recovery = round(mean(recovery < baseline) * 100, 1)
  )
```

# Establishing what "normal" looks like

A count of pixels that darkened is only interpretable against a reference. The intuitive reference is 50% — a coin flip — but that turns out to be wrong here.

Because radiance is right-skewed, the baseline *mean* sits above the baseline *median*. An entirely ordinary night therefore falls below that mean more often than half the time, purely as an artifact of the comparison. The correct reference has to be measured, not assumed.

```{r}
baseline_vec <- values(baseline_mean)[, 1]

darkened_one <- function(stack, period_label) {
  vals <- values(stack)          # extract once, not once per layer
  map_dfr(seq_len(nlyr(stack)), function(i) {
    tibble(baseline = baseline_vec, night = vals[, i]) |>
      filter(!is.na(baseline), !is.na(night)) |>
      summarize(
        date         = as.Date(names(stack)[i]),
        period       = period_label,
        n_darkened   = sum(night < baseline),
        n_total      = n(),
        pct_darkened = round(mean(night < baseline) * 100, 1)
      )
  })
}

darkened_by_night <- bind_rows(
  darkened_one(storm_stack,    "Storm"),
  darkened_one(recovery_stack, "Recovery")
) |>
  mutate(period = factor(period, levels = c("Storm", "Recovery")))

print(darkened_by_night)
```

```{r}
ggplot(darkened_by_night, aes(x = as.Date(date), y = pct_darkened, fill = period)) +
  geom_col() +
  scale_fill_manual(values = c("Storm" = "firebrick", "Recovery" = "steelblue")) +
  labs(title = "Share of Oahu pixels darker than baseline, by night",
       x = NULL, y = "% of pixels darkened", fill = NULL) +
  theme_minimal()
```

The eight event nights averaged **69%**, and every one of them exceeded the baseline average of 57%. The shift is real and consistent, but at roughly eleven percentage points it is modest — considerably smaller than a quarter of the grid going dark would suggest.

## Darkening by baseline brightness

```{r brightness-bins}
paired_values |>
  mutate(brightness_bin = cut(baseline, breaks = c(0, 5, 25, 75, Inf),
                               labels = c("Very dim", "Dim", "Medium", "Bright"))) |>
  group_by(brightness_bin) |>
  summarize(n = n(), pct_darkened = round(mean(storm < baseline) * 100, 1))
```

## Where Oʻahu's light actually is

Before examining change, it is worth establishing where the island has light to lose. Radiance is concentrated almost entirely along the southern urban corridor from Kapolei through Pearl City to East Honolulu. The Koʻolau and Waiʻanae ranges emit essentially nothing.

```{r brightness-bin-raster}
brightness_bin_raster <- classify(
  baseline_mean,
  rcl = matrix(c(
    0, 5, 1,
    5, 25, 2,
    25, 75, 3,
    75, Inf, 4
  ), ncol = 3, byrow = TRUE)
)

levels(brightness_bin_raster) <- data.frame(
  value = 1:4,
  label = c("Very dim", "Dim", "Medium", "Bright")
)

ggplot() +
  geom_spatraster(data = brightness_bin_raster) +
  scale_fill_manual(
    values = c("Very dim" = "grey85", "Dim" = "gold", "Medium" = "darkorange", "Bright" = "firebrick"),
    na.value = "transparent",
    name = "Baseline brightness"
  ) +
  labs(title = "Oahu: Brightness Bins",
       subtitle = "Very dim (baseline < 5) treated as noise floor") +
  coord_sf() +
  theme_minimal()
```

Pixels with a baseline radiance below 5 have essentially no light to lose, and including them would dilute every subsequent statistic with noise. That threshold is applied from here onward, leaving roughly 2,300 lit pixels.

```{r meaningful-mask}
meaningful_mask <- baseline_mean >= 5

baseline_meaningful <- mask(baseline_mean, meaningful_mask, maskvalues = FALSE)
storm_meaningful <- mask(storm_stack, meaningful_mask, maskvalues = FALSE)
recovery_meaningful <- mask(recovery_stack, meaningful_mask, maskvalues = FALSE)#Rec

abs_change_meaningful <- c(storm_meaningful, recovery_meaningful) - baseline_meaningful
names(abs_change_meaningful) <- c(names(storm_stack), names(recovery_stack))

pct_change_meaningful <- abs_change_meaningful / baseline_meaningful * 100
names(pct_change_meaningful) <- names(abs_change_meaningful)
```

## How much each pixel normally varies

Absolute and percent change both treat a pixel's baseline as a single fixed number. In reality some pixels are stable night to night while others fluctuate considerably with nothing unusual happening. A z-score accounts for this by scaling each change against that pixel's own typical variability.

```{r variability}
r_baseline_oahu <- crop(r_baseline, oahu_boundary) |> mask(oahu_boundary)
baseline_sd_oahu <- app(r_baseline_oahu, sd, na.rm = TRUE)

sd_values <- tibble(sd = values(baseline_sd_oahu)[, 1]) |> filter(!is.na(sd))

ggplot(sd_values, aes(x = sd)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  labs(title = "Oahu: Distribution of Per-Pixel Baseline Variability",
       x = "Standard deviation (baseline radiance)", y = "Number of pixels") +
  theme_minimal()

sd_vs_mean <- tibble(
  mean = values(baseline_mean)[, 1],
  sd = values(baseline_sd_oahu)[, 1]
) |>
  filter(!is.na(mean), !is.na(sd))

ggplot(sd_vs_mean, aes(x = mean, y = sd)) +
  geom_point(alpha = 0.15, size = 0.5) +
  labs(title = "Oahu: Baseline Mean vs. Baseline Variability, Per Pixel",
       x = "Mean baseline radiance", y = "SD of baseline radiance") +
  theme_minimal()

cv_oahu <- baseline_sd_oahu / baseline_mean
cv_values <- tibble(cv = values(cv_oahu)[, 1]) |> filter(!is.na(cv), is.finite(cv))

ggplot(cv_values, aes(x = cv)) +
  geom_histogram(bins = 60, fill = "darkorange", color = "white") +
  labs(title = "Oahu: Coefficient of Variation, Per Pixel",
       x = "CV (SD / mean baseline radiance)", y = "Number of pixels") +
  theme_minimal()
```

```{r z-score}
sd_threshold_oahu <- quantile(values(baseline_sd_oahu), 0.1, na.rm = TRUE)
baseline_sd_oahu_masked <- baseline_sd_oahu
baseline_sd_oahu_masked[baseline_sd_oahu_masked < sd_threshold_oahu] <- NA

event_stack <- c(storm_stack, recovery_stack)
event_meaningful <- mask(event_stack, meaningful_mask, maskvalues = FALSE)
names(event_meaningful) <- names(event_stack)

z_score_meaningful <- (event_meaningful - baseline_meaningful) / baseline_sd_oahu_masked
names(z_score_meaningful) <- names(event_stack)

print(minmax(z_score_meaningful))
```

## Choosing a change metric

Absolute change, percent change, and z-score answer different questions. The choice is not cosmetic: it determines which parts of the island a map draws the eye toward.

```{r metrics-table}
#| echo: false

metrics_table <- tribble(
  ~Metric, ~Formula, ~Question, ~Strengths, ~Weaknesses, ~Example,

  "Absolute change",
  "storm − baseline",
  "How much raw light was lost at this exact spot?",
  "Simple, intuitive units. No risk of blowing up near zero. Good for spotting the single brightest, most impactful drops (e.g., dense urban cores).",
  "Naturally favors bright pixels — a big drop in a bright area looks the same as a small drop in a dim one, even if the dim one lost nearly all its light.",
  "Baseline 100 → Storm 40: absolute change = −60. Baseline 8 → Storm 3: absolute change = −5. The bright pixel dominates any map or ranking, even though the dim pixel lost a similar share of its light.",

  "Percent change",
  "(storm − baseline) / baseline × 100",
  "What share of this pixel's normal light was lost?",
  "Scales fairly across pixels of different brightness. A quiet neighborhood and a bright one can be compared on the same footing.",
  "Explodes toward extreme, meaningless values when baseline is near zero — a tiny absolute change can register as thousands of percent. Requires masking out very dim / unpopulated pixels first.",
  "Baseline 100 → Storm 40: % change = −60%. Baseline 8 → Storm 3: % change = −62.5%. Now genuinely comparable — both pixels lost roughly the same share of their light.",

  "Z-score",
  "(storm − baseline mean) / baseline SD",
  "How unusual is this storm-night value, given how much this specific pixel normally fluctuates?",
  "Accounts for each pixel's own natural volatility. Surfaces real change in quiet, stable areas that a small absolute or percent change might otherwise hide. Best tool for finding locally significant events.",
  "Needs several baseline nights to estimate SD reliably. Breaks down (divides by ~zero) for perfectly flat pixels — needs the same near-zero masking as percent change. A modest z-score doesn't always mean a modest real-world impact if the pixel is naturally very volatile.",
  "Pixel A: baseline mean 8, SD 1. Storm value 3 → z = (3−8)/1 = −5. A large, unusual drop. Pixel B: baseline mean 100, SD 15. Storm value 85 → z = (85−100)/15 ≈ −1. A smaller, fairly ordinary night-to-night swing for that pixel — even though its absolute drop (−15) is three times larger than Pixel A's.",

  "Coefficient of variation (CV)",
  "baseline SD / baseline mean",
  "How naturally volatile is this pixel, independent of how bright it is?",
  "Distinguishes real relative instability from brightness alone. Useful diagnostic for deciding how much to trust a z-score at a given pixel.",
  "Not itself a change metric — describes baseline behavior only, says nothing about the storm night.",
  "Pixel with baseline mean 8, SD 1: CV = 0.13 (very stable). Pixel with baseline mean 100, SD 40: CV = 0.40 (much more erratic, relatively speaking). A z-score from the second pixel deserves more skepticism than one from the first."
)

metrics_table |>
  gt() |>
  tab_header(
    title = "Comparing Pixel-Level Change Metrics",
    subtitle = "Nighttime radiance, baseline vs. storm night"
  ) |>
  cols_label(
    Metric = "Metric",
    Formula = "Formula",
    Question = "Question it answers",
    Strengths = "Strengths",
    Weaknesses = "Weaknesses",
    Example = "Worked example"
  ) |>
  cols_width(
    Metric ~ px(140),
    Formula ~ px(160),
    Question ~ px(180),
    Strengths ~ px(220),
    Weaknesses ~ px(220),
    Example ~ px(260)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) |>
  tab_style(
    style = cell_text(weight = "bold", color = "#8B0000"),
    locations = cells_body(columns = Metric)
  ) |>
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = Formula)
  ) |>
  tab_style(
    style = cell_text(size = px(13)),
    locations = cells_body(columns = c(Question, Strengths, Weaknesses, Example))
  ) |>
  opt_row_striping() |>
  opt_table_font(font = "Helvetica") |>
  tab_options(
    table.font.size = px(14),
    heading.title.font.size = px(20),
    heading.subtitle.font.size = px(14),
    column_labels.background.color = "#f5f5f5",
    table.border.top.color = "#333333",
    table.border.bottom.color = "#333333",
    row.striping.background_color = "#fafafa"
  ) |>
  tab_footnote(
    footnote = "Baseline mean/SD computed across 7 pre-storm nights (Aug 3-6 returned no imagery); each event night compared individually.",
    locations = cells_title(groups = "subtitle")
  )
```

## Which metric this analysis relies on, and why

**Z-score drives the community rankings and the primary maps.** With a median baseline standard deviation of 0.57 radiance units, Oʻahu's pixels are volatile enough that raw differences are hard to interpret. The z-score is the only metric that scales each change against the variability of the pixel it came from.

**Percent change is reported for comparability across neighborhoods.** Oʻahu's communities differ enormously in baseline brightness, and percent change places a quiet residential area and a dense urban core on the same footing.

**Absolute change is presented first because its units are directly interpretable**, and because the contrast between the absolute and z-score maps demonstrates the weakness of relying on magnitude alone.

# Maps: change in radiance

## Absolute change, night by night

```{r map-abs-change}
#| fig-height: 8
ggplot() +
  geom_spatraster(data = abs_change_meaningful) +
  facet_wrap(~lyr, ncol = 4) +
  scale_fill_gradient2(
    low = "darkred", mid = "white", high = "darkgreen", midpoint = 0,
    na.value = "grey90", name = "Absolute\nchange"
  ) +
  labs(title = "Oʻahu: Absolute Radiance Change vs. Baseline",
       subtitle = "Red = darker than normal. Aug 14-17 storm, Aug 18-21 recovery. Very dim pixels excluded.") +
  coord_sf() +
  theme_minimal() +
  theme(axis.text = element_blank())
```

August 15 shows the strongest darkening, concentrated along the Honolulu urban corridor. Most of the island appears near-white, which illustrates the weakness described above: absolute change reveals only where there was a large amount of light to lose in the first place.

**August 16 is nearly blank, despite that being the night outages peaked.** This is the clearest visible symptom of the gap-filling problem flagged earlier — on the cloudiest night of the series, the product appears to be supplying modelled values rather than observations.

## Interactive maps: August 15

Comparing the same night under two metrics makes the difference concrete. Absolute change places nearly all of its color in one bright area near Waikīkī. The z-score map spreads the signal across the entire developed corridor, because it divides by each pixel's own variability rather than rewarding pixels for having been bright to begin with.

```{r map-abs-change-tmap}
tmap_mode("view")

tm_shape(abs_change_meaningful[["2026-08-15"]]) +
  tm_raster(
    col.scale  = tm_scale_continuous(midpoint = 0, values = "rd_yl_gn"),
    col.legend = tm_legend(title = "Absolute change, Aug 15")
  ) +
  tm_basemap("CartoDB.DarkMatter")
```

```{r map-zscore-tmap}
tmap_mode("view")

tm_shape(z_score_meaningful[["2026-08-15"]]) +
  tm_raster(
    col.scale  = tm_scale_continuous(midpoint = 0, values = "rd_yl_gn"),
    col.legend = tm_legend(title = "Z-score, Aug 15")
  ) +
  tm_basemap("CartoDB.DarkMatter")
```

## Z-score by community

Aggregating the z-score up to named communities (Census-designated places) makes it possible to compare specific neighborhoods against each other directly, rather than picking individual points by hand.

```{r cdp-boundaries}
oahu_places <- places(state = "HI", cb = TRUE, year = 2022) |>
  st_transform(4326) |>
  st_filter(oahu_boundary, .predicate = st_intersects)


```

# Z-score by community

Aggregating z-scores to named communities (Census-designated places) allows specific neighborhoods to be compared directly, rather than selecting individual points by hand. This is where the analysis produces its most convincing result.

```{r}
place_z <- map_dfr(names(z_score_meaningful), function(d) {
  tibble(
    NAME    = oahu_places$NAME,
    date    = as.Date(d),
    z_score = exact_extract(z_score_meaningful[[d]], oahu_places,
                            fun = "mean", progress = FALSE)
  )
})

oahu_places_z <- oahu_places |>
  select(NAME) |>
  left_join(place_z, by = "NAME")

ggplot(oahu_places_z) +
  geom_sf(aes(fill = z_score), color = "white", linewidth = 0.1) +
  facet_wrap(~date, ncol = 4) +
  scale_fill_gradient2(
    low = "darkred", mid = "white", high = "darkgreen", midpoint = 0,
    na.value = "grey85", name = "Mean\nz-score"
  ) +
  labs(title = "Oʻahu: Mean Radiance Z-Score by Community (CDP)",
       subtitle = "Aug 14-17 storm, Aug 18-21 recovery") +
  theme_minimal() +
  theme(axis.text = element_blank())
```

```{r place-zscore-table}
place_z |>
  filter(date <= as.Date("2026-08-17")) |>
  group_by(NAME) |>
  summarize(worst_z = min(z_score, na.rm = TRUE), .groups = "drop") |>
  filter(is.finite(worst_z)) |>
  arrange(worst_z)
```

::: callout-tip
## Independent agreement with the utility's own report

The hardest-hit communities by z-score are Ahuimanu (−5.28), Waialua (−3.15), Ewa Villages (−2.63), Maili (−2.28), East Honolulu (−2.19), West Loch Estate (−1.95), and Kahuku (−1.84).

Hawaiian Electric independently described the outage as concentrated in **Leeward Oʻahu, Honolulu, and Windward Oʻahu**. Ewa Villages, Maili, and West Loch Estate are leeward. Ahuimanu and Kahuku are windward. East Honolulu is Honolulu.

The satellite was given no information about where to look. It surfaced the same three regions the utility reported. This geographic agreement is the strongest evidence in the analysis that the measured darkening reflects power loss rather than noise or cloud.
:::
