---
title: "Soalr_detection_Rei_Miyashiro"
format: html
editor: visual
---

## Package

```{r}
library(tidyverse)
library(here)
library(sf)
library(mapview)
library(tidycensus)
```

## Data

```{r}
oahu_solar <- st_read(here("data/oahu_solar_dedup_simple.geojson"))
```

## Mapview

```{r}
mapview(oahu_solar)
```

```{r}
oahu_median_household_income <- get_acs(geography = 'tract',
                                        variables = 'B19013_001',
                                        state = 'HI',
                                        county = 'Honolulu',
                                        geometry = TRUE,
                                        year = 2022)
```

```{r}
ggplot(oahu_median_household_income, aes(x = moe, y = estimate)) +
  geom_point() +
  geom_smooth(method = 'lm')
```

\

- package is sf, language R

- head() of two datasets

- please the identify the best function(s) from sf

- Goal is to answer the question: "Is there a relationship/correlation between median income and solar infrastructure at the census tract level"?.

- one census tract example

- i got our medium income data from tidycensus

### 1. Clean and reproject

```{r}
# 1. Clean and reproject the solar layer
oahu_solar_utm <- oahu_solar |>
  filter(!st_is_empty(geometry)) |>   # drop the 68 empty geometries
  st_make_valid() |>                  # fix any self-intersecting polygons
  st_transform(26904)                 # NAD83 / UTM Zone 4N (units = meters)

# 2. Reproject the census tract layer to the same CRS
oahu_income_utm <- oahu_median_household_income |>
  st_transform(26904) |>
  st_make_valid()

# 3. Sanity checks
st_crs(oahu_solar_utm) == st_crs(oahu_income_utm)  # should be TRUE
nrow(oahu_solar_utm)                               # should be 88663 - 68 = 88595
```

### 2. Convert panels to points and join them to tracts

```{r}
# 1. Reduce each solar polygon to one point guaranteed to fall inside it
solar_points <- oahu_solar_utm |>
  st_point_on_surface()

# 2. Spatially join each point to the census tract it falls within
solar_tract <- st_join(
  solar_points,
  oahu_income_utm |> select(GEOID),
  join = st_within
)

# 3. Sanity checks
nrow(solar_tract)              # should still be 88595 (no duplicated panels)
sum(is.na(solar_tract$GEOID))  # panels that didn't land in any tract
head(solar_tract)
```

### 3. Sum solar area per tract and join back to all tracts

```{r}
# 1. Sum solar area per tract (as a plain table, not spatial)
solar_by_tract <- solar_tract |>
  st_drop_geometry() |>              # drop point geometry, keep attributes
  filter(!is.na(GEOID)) |>           # set aside the 170 unmatched panels
  group_by(GEOID) |>
  summarise(
    n_panels      = n(),
    solar_area_m2 = sum(area_m2, na.rm = TRUE)
  )

# 2. Attach the totals back onto ALL census tracts
tract_solar <- oahu_income_utm |>
  left_join(solar_by_tract, by = "GEOID") |>
  mutate(
    n_panels      = replace_na(n_panels, 0L),
    solar_area_m2 = replace_na(solar_area_m2, 0)
  )

# 3. Sanity checks
nrow(tract_solar) == nrow(oahu_income_utm)       # no tracts lost or duplicated
sum(tract_solar$solar_area_m2)                   # total solar area assigned
sum(solar_tract$area_m2[!is.na(solar_tract$GEOID)], na.rm = TRUE)  # should match
sum(tract_solar$n_panels == 0)                   # tracts with no detected solar
sum(is.na(tract_solar$estimate))                 # tracts with no income estimate
```

### 4. normalize solar area and build the analysis dataset

```{r}
# 1. Get household counts for each tract (same ACS year as income)
oahu_households <- get_acs(geography = "tract",
                           variables = "B11001_001",   # total households
                           state     = "HI",
                           county    = "Honolulu",
                           year      = 2022,
                           geometry  = FALSE) |>
  select(GEOID, households = estimate)

# 2. Add tract area, household counts, and normalized solar measures
tract_solar <- tract_solar |>
  mutate(tract_area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  left_join(oahu_households, by = "GEOID") |>
  mutate(
    solar_m2_per_hh  = solar_area_m2 / households,
    solar_m2_per_km2 = solar_area_m2 / tract_area_km2
  )

# 3. Keep only tracts usable for the income comparison
analysis_df <- tract_solar |>
  filter(!is.na(estimate), households > 0) |>
  rename(median_income = estimate)

# 4. Sanity checks
nrow(analysis_df)
analysis_df |>
  st_drop_geometry() |>
  select(median_income, households, tract_area_km2,
         solar_area_m2, solar_m2_per_hh, solar_m2_per_km2) |>
  summary()
```

```{r}
# 1. Look at the most extreme tracts
analysis_df |>
  st_drop_geometry() |>
  slice_max(solar_m2_per_hh, n = 5) |>
  select(GEOID, NAME, median_income, households,
         solar_area_m2, solar_m2_per_hh)

# 2. Scatterplot: income vs. solar per household
ggplot(analysis_df, aes(x = median_income, y = solar_m2_per_hh)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  scale_x_continuous(labels = scales::label_dollar()) +
  scale_y_log10() +
  labs(
    title = "Median household income vs. solar panel area per household",
    subtitle = "Oʻahu census tracts, ACS 2022",
    x = "Median household income",
    y = "Solar panel area per household (m², log scale)"
  ) +
  theme_minimal()

# 3. Spearman rank correlation for both solar measures
cor.test(analysis_df$median_income, analysis_df$solar_m2_per_hh,
         method = "spearman", exact = FALSE)

cor.test(analysis_df$median_income, analysis_df$solar_m2_per_km2,
         method = "spearman", exact = FALSE) 
```
